Hunting experience shapes individual foraging specialization and predator-prey interactions in an online videogame

### 2023-06-02 manuscript/Chapter2-manuscript.Rmd
updated methods MS and figure captions

The videogame company provided data that spanned a period of 6 months of gameplay recorded for every player from `r toString(format(min(dat$date)))` to `r toString(format(max(dat$date)))`. We only analyzed matches where players did not know each other (i.e. "Online" mode). We filtered any matches where players were inactive, such as when average distances traveled per second (i.e. speed) were equal to, or very close to, zero. Moreover, we used our knowledge of the game to remove any matches where players were potentially hacking, or not playing the game as intended. We then sampled players that played 300 matches or more, and monitored all their matches from the first to a maximum of 500 matches.

Our population consists of `r toString(format(length(unique(dat$predator_id))))` predator players with a total record of `r toString(formatC(length(unique(dat$match_id)), big.mark = " "))` matches. The predator-players' experience varied between `r toString(format(min(dat$total_xp_pred)))` and `r toString(format(max(dat$total_xp_pred)))` matches played. These matches lasted between `r toString(format(round(min(dat$game_duration / 60))), digits = 2)` and `r toString(format(round(max(dat$game_duration / 60))), digits = 2)` min (mean = `r toString(format(round(mean(dat$game_duration / 60))), digits = 2)` min). The following information is collected and reported for every match : the player's anonymous ID, its avatar (i.e. the predator character chosen with its specific skill-gameplay mechanics), the game environment, the predator-player's experience along with its speed, and the average speed of the group of prey it encountered.

We analyzed the predator's average speed and the average speed of the prey group encountered by the predator. The predator's average speed is measured as the average distance traveled per second during a match (mean= `r toString(format(round(mean(dat$pred_speed), digits = 2), nsmall = 2))` ± `r toString(format(round(sd(dat$pred_speed), digits = 2), nsmall = 2))` m/s). We measured the preys' average speed as the average travel speed of the four individual prey within a match (mean = `r toString(format(round(mean(dat$prey_avg_speed, na.rm = T), digits = 2), nsmall = 2))` ± `r toString(format(round(sd(dat$prey_avg_speed, na.rm = T), digits = 2), nsmall = 2))` m/s). We defined hunting success as the number of prey consumed during the match (min = `r toString(format(min(dat$hunting_success)))`, max = `r toString(format(max(dat$hunting_success)))`). Lastly, we defined the predator's cumulative experience as the number of matches played prior to the match being monitored. For example, the first match of a player would have a cumulative experience value of 0, while the tenth match would have a value of 9. We also categorized predators for each match based on their cumulated experience. We labeled predators as novices for matches where they had cumulated less than 100 matches, intermediate for matches where they had cumulated between 100 and 299 matches, and advanced for matches where they had cumulated more than 299 matches (max 499). Since our goal was to monitor predator players throughout their experience and that they all played at least 300 matches, they all appeared in each of the three experience categories.

We recognize that we could have biased our analyses by retaining only individuals who played for at least 300 matches. For example, these individuals might be experienced videogame players and could thus already be playing like experts in their first matches in *DBD*. To verify that our sample was not biased, we compared a random sample of players that played either 20 to 50 matches, 51 to 100 matches, or 101 to 300 matches during the same timeframe as our sampled population. We then took the first 20 matches played by these players, including those from our sampled population, and compared their average speed using a Bayesian hierarchical linear model. We found that all four groups had similar average speeds as predators (Appendix 2: Table S1-S2), which indicates an absence of bias due to data sampling.

@@

We tested how prey influenced the development of predator expertise by computing five Bayesian generalized additive mixed models (GAMM) with thin-plate regression splines, all of which estimated the relationship between hunting success (i.e. number of prey consumed) and the predators' cumulative experience (i.e. number of matches played before the current match). We parameterized the models following the method of @Pedersen.etal2019. The first model was the simplest, with a common global smoothing function and random intercepts for the predator ID. In this model, we assume that predators have the same development of expertise, with the model estimating a trend for the average individual (i.e. global smoother). For the second model, we included varying individual smoothers for the predator ID. Here, we assume that individual predators share a similar relationship between success and experience, but that this relationship can vary among them (e.g. predator 1 has a steeper curve than predator 2). This enabled us to test whether predators differed in the development of their expertise. In the third model, we kept the individual smoothers for the predators, but removed the global smoother. This model assumes that predators do not share a common relationship between success and experience. The fourth and fifth models were a reproduction of the second and third models respectively, where we included the prey group's average speed to assess its effect on the relationship between success and experience. We included the standardized match duration and average prey rank (a proxy for prey skill) as covariates in all five models. The ranking system in *DBD* was implemented by the company to pair players in a match based on their skill (https://deadbydaylight.fandom.com/wiki/Rank), and failing to account for it would prevent us from detecting a change in the predator's foraging success with experience.

@@

We tested whether foraging behaviour changed with experience, and compared the success of predators along the flexible-specialist hunter continuum, using a multivariate double-hierarchical generalized linear model (MDHGLM) [@Lee.Nelder2006;@Cleasby.etal2015;@Mitchell.etal2016a]. We fitted the model on the predator's speed ($Y_{1}$), the prey's speed ($Y_{2}$), and the predator's hunting success ($Y_{3}$), using a Gaussian distribution family for the predator and the prey's speed, and a beta-binomial distribution family for the predator's success.

The model simultaneously estimated three distinct components for each experience level and each trait. The first component is among individual variation in mean speed and success (the mean part of the model), allowing to test for individual differences in tactic use (i.e. along the slow-fast continuum), in the mean speed of the prey encountered, and in mean prey consumption. The second component is among individual variation in intra-individual behavioural variation (IIV) (the dispersion part of the model), allowing us to test for individual differences in foraging specialization and in how (dis)similar were the groups of prey encountered (i.e. if resource heterogeneity was higher for some individuals compared to others). The third component consists of pairwise comparisons between the individual identity random effect standard deviations on the mean and dispersion part of the model within/across traits within experience levels, and within/across traits across experience levels, resulting in a 15x15 (co)variance matrix. This allows us to assess whether predators that are on average slower/faster are more specialized or flexible, whether predators that adopted a slower/faster (or more specialized/flexible) tactic encountered prey that were faster/slower, and whether slower/faster predators had lower/greater hunting success.

We controlled for the standardized average rank of the prey group for both the predator and the prey's speed on both the mean and the dispersion parts of the model. We included game environment and predator avatar random effects on the mean part of the model exclusively that were trait and experience specific (i.e. no covariance structure). We included individual predator identity random effects on the mean and dispersion parts. For hunting success, we modeled the mean part only and controlled for the standardized mean prey rank and game duration, and included an individual identity random effect.

@@

eqs_num <- c(
    "(1.1)",
    "(1.2)",
    "(2.1)",
    "(2.2)",
    "(3)"
)

@@

# Customize table
my_theme <- function(x, ...) {
  x <- colformat_double(x, big.mark = " ",
                        decimal.mark = ".",
                        digits = 2)
  x <- border_remove(x)
  x <- delete_part(x, part = "header")
  x <- align(x, j = 1, align = "left", part = "body")
  x <- align(x, j = 2, align = "right", part = "body")
  x <- font(x, fontname = "Times New Roman", part = "body")
  x <- fontsize(x, size = 12, part = "body")
  x <- width(x, j = 1, width = 15.25, unit = "cm")
  x <- width(x, j = 2, width = 1.26, unit = "cm")
}

@@

where $j$ is the index of the experience levels. The model is fitted on $n$ experience levels ($n = 3$) denoted by $T_{j}$, where $j$ ranges from 1 to $n$, being the novice, intermediate, and advanced levels. The $b$ terms are vectors of fixed effects with their design matrices $X$. The terms $en$, $av$, and $id$ are the vectors of random environment, random avatar, and random individual identity effects associated with their design matrices $M$, $N$, and $Z$. The $d$ subscript indicates that a term is from the dispersion part of the model. On the mean part for all traits and experience levels, we assumed that the game environment and avatar random effects followed a Gaussian distribution with estimated standard deviation ($en \sim N(0,\ I\ \sigma_{en})$ and $av \sim N(0,\ I\ \sigma_{av})$). $I$ is the identity matrix, with $I\sigma$ indicating that the random effects are independently and identically distributed. The individual identity random effects on both the mean and dispersion parts across all response variables and experience levels are indexed in vector $u$. The individual identity random effects contained in vector $u$ follow a multivariate Gaussian distribution $u \sim MVN(0,\ \Sigma)$ where $\Sigma$ is a 15x15 (co)variance matrix of the individual identity standard deviations $\sigma_{id_i}$ across all experience levels and response variables.

@@

Individual predators differed slightly in their average movement speed across experience levels (i.e. predator ID (mean); Figure 2 and Appendix S1: Table S2). However, there was strong evidence that they displayed differences in foraging specialization, with a 0.1 unit increase in standard deviation from novice to advanced, indicating that differences increased slightly (i.e. predator ID (sigma) is higher at the advanced level; Figure 2 and Appendix S1: Table S2). Moreover, some individual predators switched from a flexible to a specialized hunting tactic, and vice-versa, as they gained experience (Figure 3). Importantly, `r toString(scales::percent(percent_tab2[1,]$percentage, accuracy = 1))` of the predator population displayed a large change in foraging specialization/flexibility with experience (i.e. >0.2 change in standard deviation), `r toString(scales::percent(percent_tab2[2,]$percentage, accuracy = 1))` displayed lower changes (i.e. >0.05 and <0.2 change in standard deviation), while `r toString(scales::percent(percent_tab2[3,]$percentage, accuracy = 1))`remained relatively stable (i.e. <0.05 change in standard deviation).

Contrary to predictions, there was strong evidence that predators did not differ in the speed of the prey groups that they encountered on average across experience (i.e. predator ID (mean); Figure 2 and Appendix S1: Table S2). However, individuals differed in the heterogeneity of the groups of prey encountered across the study period (i.e. predator ID (sigma); Appendix S1: Table S2). Hence, some predators encountered prey groups with similar average speeds, while others encountered prey groups with varying average speeds throughout the study period. The differences in the prey encountered among predators increased only slightly with experience (i.e. predator ID (sigma); Figure 2).

@@

Surprisingly, the effect of experience on foraging behaviour was much more subtle than we expected. For instance, the predator population maintained their speed and degree of specialization throughout the study period, and individual differences in mean speed did not change with experience. However, differences among predators in foraging specialisation did increase slightly with experience, with some individuals becoming increasingly flexible. Interestingly, the behavioural response (i.e. specialization) to experience by individual predators differed. For instance, 44% of the population did not alter their degree of specialization with experience (i.e. remaining either faster and more specialized, or slower and more flexible), while 56% of the population displayed a switch from a specialized to a flexible tactic or vice-versa with experience. These changes were accompanied by predators increasingly matching their specialization to the mean speed of the prey groups across experience, instead of the variability between encounters. This suggests that it may be easier for predators to specialize on their prey by relying on mean cues (i.e. mean behaviour) instead of relying solely on resource heterogeneity [@Weimerskirch.etal2005;@Weimerskirch2007;@Woo.etal2008;@Phillips.etal2017].

@@

**Figure 1.** Median posterior predictions of the development of predator hunting expertise. The predators' hunting success (i.e. the probability of consuming the four prey) is on the y axis, and the predators' cumulative experience (i.e. the number of matches played prior to each observation) is on the x axis. Panels A and C show among individual differences in the development of expertise, while panels B and D show the development of expertise for the average individual. (A-B) GAMM where we control for the average prey rank (C-D) GAMM where we control for the average prey rank and the average speed of the prey group.

**Figure 2.** Median posterior differences and HPD intervals comparing the parameter values predicted by the MDHGLM among all predator experience levels. The parameter being compared is on the y axis, and the difference in parameter values between two experience levels is on the x axis. The pairwise comparisons between experience levels are displayed across the three panels A, B, and C. The population mean corresponds to the population's mean speed, mean speed of the prey encountered, and mean success. The population variation corresponds to the population's foraging specialization and variation in speed of the prey encountered. The individual variation in means corresponds to individual differences in mean speed, while individual variation in IIV indicates individual differences in foraging specialization and in the variation of the groups of prey encountered.

>>>>

The videogame company provided data that spanned a period of 6 months of gameplay recorded for every player from `r toString(format(min(dat$date)))` to `r toString(format(max(dat$date)))`. We only analyzed matches where players did not know each other (i.e. "Online" mode). We filtered any matches where players were inactive, such as when mean distances traveled per second (i.e. speed) were equal to, or very close to, zero. Moreover, we used our knowledge of the game to remove any matches where players were potentially hacking, or not playing the game as intended. We then sampled players that played 300 matches or more, and monitored all their matches from the first to a maximum of 500 matches.

Our population consists of `r toString(format(length(unique(dat$predator_id))))` predator players with a total record of `r toString(formatC(length(unique(dat$match_id)), big.mark = " "))` matches. The predator-players' experience varied between `r toString(format(min(dat$total_xp_pred)))` and `r toString(format(max(dat$total_xp_pred)))` matches played. These matches lasted between `r toString(format(round(min(dat$game_duration / 60))), digits = 2)` and `r toString(format(round(max(dat$game_duration / 60))), digits = 2)` min (mean = `r toString(format(round(mean(dat$game_duration / 60))), digits = 2)` min). The following information is collected and reported for every match : the player's anonymous ID, its avatar (i.e. the predator character chosen with its specific skill-gameplay mechanics), the game environment, the predator-player's experience along with its speed, the mean speed of the groups of prey the player encountered, and their mean rank (a proxy for prey skill). The ranking system in *DBD* was implemented by the company to pair players in a match based on their skill (https://deadbydaylight.fandom.com/wiki/Rank), and failing to account for it would prevent us from detecting a change in the predator's foraging success with experience.

We analyzed the predator's mean speed and the mean speed of the prey group encountered by the predator. The predator's mean speed is measured as the mean distance traveled per second during a match (mean= `r toString(format(round(mean(dat$pred_speed), digits = 2), nsmall = 2))` ± `r toString(format(round(sd(dat$pred_speed), digits = 2), nsmall = 2))` m/s). We measured the preys' speed as the mean travel speed of the four individual prey in a match (mean = `r toString(format(round(mean(dat$prey_avg_speed, na.rm = T), digits = 2), nsmall = 2))` ± `r toString(format(round(sd(dat$prey_avg_speed, na.rm = T), digits = 2), nsmall = 2))` m/s). We defined hunting success as the number of prey consumed during the match (min = `r toString(format(min(dat$hunting_success)))`, max = `r toString(format(max(dat$hunting_success)))`). Lastly, we defined the predator's cumulative experience as the number of matches played prior to the match being monitored. For example, the first match of a player would have a cumulative experience value of 0, while the tenth match would have a value of 9. We also categorized predators for each match based on their cumulated experience. We labeled predators as novices for matches where they had cumulated less than 100 matches, intermediate for matches where they had cumulated between 100 and 299 matches, and advanced for matches where they had cumulated more than 299 matches (max 499). Since our goal was to monitor predator players throughout their experience and that they all played at least 300 matches, they all appeared in each of the three experience categories.

We recognize that we could have biased our analyses by retaining only individuals who played for at least 300 matches. For example, these individuals might be experienced videogame players and could thus already be playing like experts in their first matches in *DBD*. To verify that our sample was not biased, we compared a random sample of players that played either 20 to 50 matches, 51 to 100 matches, or 101 to 300 matches during the same timeframe as our sampled population. We then took the first 20 matches played by these players, including those from our sampled population, and compared their mean speed using a Bayesian hierarchical linear model. We found that all four groups had similar mean speeds as predators (Appendix 2: Table S1-S2), which indicates an absence of bias due to data sampling.

@@

We tested how prey influenced the development of predator expertise by computing five Bayesian generalized additive mixed models (GAMM) with thin-plate regression splines, all of which estimated the relationship between hunting success (i.e. number of prey consumed) and the predators' cumulative experience (i.e. number of matches played before the current match). We parameterized the models following the method of @Pedersen.etal2019. The first model was the simplest, with a common global smoothing function and random intercepts for the predator ID. In this model, we assume that predators have the same development of expertise, with the model estimating a trend for the average individual (i.e. global smoother). For the second model, we included varying individual smoothers for the predator ID. Here, we assume that individual predators share a similar relationship between success and experience, but that this relationship can vary among them (e.g. predator 1 has a steeper curve than predator 2). This enabled us to test whether predators differed in the development of their expertise. In the third model, we kept the individual smoothers for the predators, but removed the global smoother. This model assumes that predators do not share a common relationship between success and experience. The fourth and fifth models were a reproduction of the second and third models respectively, where we included the prey speed to assess its effect on the relationship between success and experience. We included the standardized match duration and prey rank as covariates in all five models.

@@

We tested whether foraging behaviour changed with experience, and compared the success of predators along the flexible-specialist hunter continuum, using a multivariate double-hierarchical generalized linear model (MDHGLM) [@Lee.Nelder2006;@Cleasby.etal2015;@Mitchell.etal2016a]. We fitted the model on both the predator ($Y_{1}$) and prey ($Y_{2}$) speeds using a Gaussian error distribution, and on the predator's hunting success ($Y_{3}$) using a beta-binomial error distribution.

The model simultaneously estimated three distinct individual components for each trait across experience levels. The first component, estimated on the mean part of the model, is among individual variation in mean speed and success. This allows to test for individual differences in tactic use (i.e. along the slow-fast continuum), in the mean speed of the prey encountered, and in mean prey consumption. The second component, estimated on the dispersion part of the model, is among individual variation in intra-individual variation (IIV). This allows to test if predators do not experience the same degree of prey heterogeneity and if they differ in individual specialization . The third component consists of pairwise comparisons between the individual identity standard deviations on the mean and dispersion part of the model among traits, among experience, and across traits and experience. This results in a 15x15 (co)variance matrix, allowing to describe the structure of the predator-prey trait interactions as well as the relationship between specialization and success across experience at the individual-level.

For the predator and the prey speed, we controlled for the standardized prey rank on both the mean and the dispersion parts of the model. We included game environment, predator avatar, and individual predator identity random effects on the mean part of the model. The dispersion part included individual predator identity random effects exclusively. For hunting success, we modeled the mean part only and controlled for the standardized prey rank and game duration, and included an individual identity random effect.

@@

eqs_num <- c(
    "(eqn. 1.1)",
    "(eqn. 1.2)",
    "(eqn. 2.1)",
    "(eqn. 2.2)",
    "(eqn. 3)"
)

@@

# Customize table
my_theme <- function(x, ...) {
  x <- colformat_double(x, big.mark = " ",
                        decimal.mark = ".",
                        digits = 2)
  x <- border_remove(x)
  x <- delete_part(x, part = "header")
  x <- align(x, j = 1, align = "left", part = "body")
  x <- align(x, j = 2, align = "right", part = "body")
  x <- font(x, fontname = "Times New Roman", part = "body")
  x <- fontsize(x, size = 12, part = "body")
  x <- width(x, j = 1, width = 13, unit = "cm")
  x <- width(x, j = 2, width = 2.26, unit = "cm")
}

@@

where $j$ is the index of the experience levels. The model is fitted on the three experience levels denoted by $T_{j}$, where $j$ ranges from 1 to 3 (i.e. novice, intermediate, advanced). The $b$ terms are vectors of fixed effects with their design matrices $X$. The terms $en$, $av$, and $id$ are the vectors of random environment, random avatar, and random individual identity effects associated with their design matrices $M$, $N$, and $Z$. The $d$ subscript indicates that a term is from the dispersion part of the model. For all experience levels of the predator and prey speed, we assumed that the game environment and avatar random effects followed a Gaussian distribution with estimated standard deviation ($en \sim N(0,\ I\ \sigma_{en})$ and $av \sim N(0,\ I\ \sigma_{av})$). $I$ is the identity matrix, with $I\sigma$ indicating that the random effects are independently and identically distributed. All the individual identity random effects (mean and dispersion) across all response variables and experience levels are indexed in vector $u$. The $u$ vector follows a multivariate Gaussian distribution $u \sim MVN(0,\ \Sigma)$ where $\Sigma$ is a 15x15 (co)variance matrix of the individual identity standard deviations $\sigma_{id_i}$ across all experience levels and response variables.

@@

Individual predators differed slightly in their mean speed across experience levels (i.e. predator ID (mean); Figure 2 and Appendix S1: Table S2). However, there was strong evidence that they displayed differences in foraging specialization, with a 0.1 unit increase in standard deviation from novice to advanced, indicating that differences increased slightly (i.e. predator ID (sigma) is higher at the advanced level; Figure 2 and Appendix S1: Table S2). Moreover, some individual predators switched from a flexible to a specialized hunting tactic, and vice-versa, as they gained experience (Figure 3). Importantly, `r toString(scales::percent(percent_tab2[1,]$percentage, accuracy = 1))` of the predator population displayed a large change in foraging specialization/flexibility with experience (i.e. >0.2 change in standard deviation), `r toString(scales::percent(percent_tab2[2,]$percentage, accuracy = 1))` displayed lower changes (i.e. >0.05 and <0.2 change in standard deviation), while `r toString(scales::percent(percent_tab2[3,]$percentage, accuracy = 1))`remained relatively stable (i.e. <0.05 change in standard deviation).

Contrary to predictions, there was strong evidence that predators did not differ in the mean speed of the prey groups that they encountered across experience (i.e. predator ID (mean); Figure 2 and Appendix S1: Table S2). However, individuals differed in the heterogeneity of the groups of prey encountered across the study period (i.e. predator ID (sigma); Appendix S1: Table S2). Hence, some predators encountered prey groups with similar mean speeds, while others encountered prey groups with varying mean speeds throughout the study period. The differences in the prey encountered among predators increased only slightly with experience (i.e. predator ID (sigma); Figure 2).

@@

Surprisingly, the effect of experience on foraging behaviour was much more subtle than we expected. For instance, the predator population maintained their speed and degree of specialization throughout the study period, and individual differences in mean speed did not change with experience. However, differences among predators in foraging specialization did increase slightly with experience, with some individuals becoming increasingly flexible. Interestingly, the behavioural response (i.e. specialization) to experience by individual predators differed. For instance, 44% of the population did not alter their degree of specialization with experience (i.e. remaining either faster and more specialized, or slower and more flexible), while 56% of the population displayed a switch from a specialized to a flexible tactic or vice-versa with experience. These changes were accompanied by predators increasingly matching their specialization to the mean speed of the prey groups across experience, instead of the variability between encounters. This suggests that it may be easier for predators to specialize on their prey by relying on mean cues (i.e. mean behaviour) instead of relying solely on resource heterogeneity [@Weimerskirch.etal2005;@Weimerskirch2007;@Woo.etal2008;@Phillips.etal2017].

*expliquer pourquoi les généralistes ont moins bien performé ici*

@@

**Figure 1.** Median posterior predictions of the development of predator hunting expertise. The predators' hunting success (i.e. the probability of consuming the four prey) is on the y axis, and the predators' cumulative experience (i.e. the number of matches played prior to each observation) is on the x axis. Panels A and C show among individual differences in the development of expertise, while panels B and D show the development of expertise for the average individual. (A-B) GAMM where we control for the prey rank (C-D) GAMM where we control for the prey rank and the speed of the prey group.
